---
title: "mixedWarpedCurves2"
output: 
  github_document:
    toc: false
    toc_depth: 1
    pandoc_args: --webtex
---

[![Travis-CI Build Status](https://travis-ci.org/eric-f/mixedWarpedCurves2.svg?branch=master)](https://travis-ci.org/eric-f/mixedWarpedCurves2) [![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/eric-f/mixedWarpedCurves2?branch=master&svg=true)](https://ci.appveyor.com/project/eric-f/mixedWarpedCurves2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.path = "man/figures/README-")
```

# Dependencies

- GNU Scientific Library (GSL)
- R packages:
    - splines
    - gtools
    - Rcpp
    - RcppArmadillo
    - RcppGSL
    - RcppProgress
    - BH

On Windows, the R package will be built using the GSL at https://github.com/rwinlib/gsl. On other platforms, you will need to have the library installed manually before this R package can be compiled.

On OSX, you can install GSL via [Homebrew](https://brew.sh). If you do not already have Homebrew, go to https://brew.sh for installation instructions. After that, run the following command in your terminal
```
brew install gsl
```

On Debian or Ubuntu, install from terminal by running
```
sudo apt-get install libgsl0-dev
```

# Installation

Once all the dependencies have been installed, run the following lines in R to install the latest version of this package:
```
library(devtools)
devtools::install_github('eric-f/mixedWarpedCurves2')
```

# Overview

The main functions in this package to fit the warped curves model via Stochastic approximation EM algorithms are `fsim()` and `fsim_unimodal()`. Control arguments for the model and algorithm can be specificy via `control_saem()`.

# Warped curves model

We model the functional data as noisy observations of latent curves, $y_i$'s, at time $t_{ij}$, as $y_{ij} = y_i(t_{ij}) + \epsilon_{ij}$.

The latent curve, $y_i(t) = f_i \circ h_i(t),$ is composed of two functions, $f_i$ and $h_i$, representing amplitude and phase variations respectively. Specifically, the amplitude function has the form: $f_i = a_{i,sh} + a_{i,sc} f(t)$, where $f$ is a common base shape whereas the warping function is continuous, strictly increasing and satisfies $h_i(0)\equiv 0$ and $h_i(1)\equiv 1$.

## Amplitude component

The base shape, $f$, is either a known function or modelled as a spline that is to be estimated. The amplitude random effect, $(a_{i,sh}, a_{i,sc})$, is Gaussian with mean $(\mu_{sh}, \mu_{sc})$ and variance-covariance matrix $\Sigma$.

The fixed base shape version is implemented in `fsim_unimodal()` where $f(t) = -4t(1-t)$.

The unknown base shape version is implemented `fsim_mixed_warped_curves()` where the base shape is modelled using a B-spline. The order and knot locations of the B-spline are specified by the arguments `f_order` and `f_knots` via `control_saem()`.

## Phase component

The warping functions are modelled using B-splines,
$$h_i(t) = \sum_{k=1}^{K_h} w_{i,k}B_k(t),$$
where $0\equiv w_{i,1} \leq \cdots \leq w_{i,K_h} \equiv 1$. The first difference of the random coefficients,
$(w_{i,2}, w_{i,3}-w_{i,2}, \ldots, 1-w_{i,K_h-1}),$
follows a Dirichlet distribution whose mean is fixed such that $E(h_i(t))=t$.

The order and knot locations of the B-splines are specified by the arguments `h_order` and `h_knots` via `control_saem()`.


# Usage
This package comes with a simulated dataset, `data_ex1`, of 50 curves observed on 101 equally spaced time points.

```{r example-1-data, fig.width=4, fig.show='hold'}
library(mixedWarpedCurves2)
library(ggplot2)
library(plyr)
library(dplyr)

## Plot base shape
data_ex1 %>%
  dplyr::filter(id=="1") %>%
  ggplot() + 
  geom_line(aes(x=warped_x, y=(y0 - sh)/sc)) + 
  labs(x="Time, t", y=expression(paste("Base shape, ", f(t))))

## Plot observed curves
ggplot(data_ex1) + 
  geom_line(aes(x=x, y=y, col=id), size=0.5, show.legend = F) + 
  labs(x=expression(paste("Time, ", t[ij])), 
       y=expression(paste("Observed curve, ", y[ij])))

## Plot underlying amplitude functions
ggplot(data_ex1) + 
  geom_line(aes(x=warped_x, y=y0, col=id), show.legend = F) + 
  labs(x="Time, t", y=expression(paste("Amplitude function, ", f[i](t))))

## Plot underlying warping functions
ggplot(data_ex1) + 
  geom_line(aes(x=x, y=warped_x, col=id), show.legend = F) + 
  labs(x="Time, t", 
       y=expression(paste("Warping function, ", h[i](t))))
```
The figures above show the base shape (top left), the observe curves (top right), the amplitude functions, $f_i$, (bottom left) and the warping functions, $h_i$, (bottom right).

Call `fsim()` to fit the model by maximum likelihood via a stochastic approximation expectation-maximization (SAEM) algorithm:
```{r example-1-saem, cache=FALSE}
out <- fsim(
  y = data_ex1$y,
  obs_time = data_ex1$x,
  curve_id = data_ex1$id,
  saem_control = control_saem(
    seed = 0,
    n_saem_iter = 1000,
    n_saem_burn = 100,
    n_mcmc_burn = 5,
    h_knots = seq(0, 1, length=5),
    f_knots = seq(0, 1, length=7)
  ))
```


The returned objects has the following components:
```{r str-out}
names(out)
```
See documentation in R for details of the output:
```
help(fsim)
```

The fitted curves and the warping functions can be extract by:
```{r example-1-output}
out_df <- ldply(out$curves, function(crv){
  data.frame(
    id = as.character(c(crv$curve_id)),
    x = c(crv$x),
    y = c(crv$y),
    warped_x = c(crv$warped_x),
    fitted_y = c(crv$fitted_y)
  )
})
```

The figures below shows the fitted curves (left) and the registered curves (right).
```{r example-1-plots, fig.width=4, fig.show='hold'}
## Plot fitted curves
ggplot(out_df) + 
  geom_line(aes(x=x, y=y, group=id), size=0.3) + 
  geom_line(aes(x=x, y=fitted_y, col=id), 
            size=0.3, show.legend = F) + 
  labs(x="Time, t", 
       y=expression(paste("Fitted curve, ", hat(y)[i](t))))

## Plot registered curves
ggplot(out_df) + 
  geom_line(aes(x=warped_x, y=y, col=id),
            show.legend = F) + 
  labs(x=expression(paste("Predicted warped time, ", hat(h)[i](t))), 
       y=expression(y[i](t)))
```

The sequences of estimated parameters are returned in the `pars_track` component. The figure below shows the estimated error variances as a simple convergence diagnostics:
```{r example-1-tracks, fig.width=12}
sigma2_track <- out$pars_track$sigma2_track
qplot(x=seq(along=sigma2_track),
      y=sigma2_track, 
      geom="line") + 
  labs(x="Iterations", 
       y=expression(paste("Estimated ", sigma[epsilon]^2))) + 
  coord_cartesian(ylim=c(sigma2_track[50], min(sigma2_track)))
  
```

# Reference
Fu, E. and Heckman, N. (2017). Model-based curve registration via stochastic approximation EM algorithm. \textit{arXiv preprint arXiv:1712.07265}
https://arxiv.org/abs/1712.07265
